Developmental trajectories of word recognition: a priming study in bilinguals

### 2023-10-30 manuscript/manuscript.qmd
Clean repo

targets::tar_load(
	c(model_fits_related,
	  model_fits_cognate,
	  # model_fits_related_time,
	  # model_fits_cognate_time,
	  # model_loo_related,
	  # model_loo_cognate,
	))

@@

@@

For instance, previous work on lexical acquisition has provided some insight into how bilinguals' vocabulary size is equivalent to monolingual's across ages, and the possible mechanisms behind this. @hoff2012 found that English-Spanish bilinguals' vocabulary in English was influenced by the words they had already acquired in Spanish. This is consistent with previous accounnts of lexical acquisition in which infants acquire new words through preferential acquisition [@hill2013, @fourtassi2020]. A monograph by @floccia2018introduction provided evidence of a language distance effect, where bilinguals learning two languiages sharing higher lexical similarity knew more words in their less dominant language than those learning two language with lower lexical similarity. Later work by @mitchell2022cognate and @garcia-castro@cognate provided evidence for a more direct facilitation effect of cognateness. Cognates are form-simlar translation equivalents. Cognateness is suggested to facilitate word acquisition through parallel activation: the co-activation of translation equivalents in speech might lead to an faster accumulation of learning instances with words, which ultimately results in an earlier age of acquisition.

It is well established that, across languages, monolingual infants start recognising their first word-forms before their first year of life [@jusczyk1999; @halle1994; @vihman2004], with the earliest age at which there is evidencer of familiar word recognition being 6 months [@bergelson2012]. Bilinguals start recognising words at similar ages. Using a Head-turn Prefernce Procedure and ERP registration,  @vihman2013 found English-Welsh bilinguals to be sensitive to the prevalence of auditorily presented words. Infants' looking times and ERP differed when presented with words frequently marked as "familiar" by caregivers of demogrtaphically equivalent peers (e.g., *nappy*, *thank you*, *bottle*, *apple*, and *naughty*), compared to when presented with rare words (e.g., *Eiffel*, *courtly*, *nettle*, *wacky*, and *juncture*).

This is also the case in the developing lexicon. There is evidence that infants start forming their first lexical representations before their first year of life. Already at six months of age, infants show evidence of word recognition.  When presented with pairs of familiar objects and hearing the label of one of them, 6-months-old infants and older show a looking preference towards the named object [@bergelson20126; @bergelson2015early]. @chow2017spoken adapted the Visual World Paradigm from @huettig2007tug to explore 24- to 30-month-old toddlers' visual fixation patterns during a word recognition when presented with phonological and semantic distractors. In each trial, the authors presented participants with four semantically and phonologically pictures. Four seconds after pictures onset, a word-form was auditorily presented. The word did not refer to any of the pictures displayed on the screen, but was phonologically related to one of them (both labels shared phonological onset), and semantically related to another one of the pictures (both referents belonged to the same taxonomic category. For instance, participants might be presented with the pictures of a sandwich, a bus, a cat, and a dress. Then they would hear the carrier phrase "Look at the **bee**!". The authors registered participants fixations to the phonological and semantic distractors, and found evidence of a preference for the phonological distractor at earlier stages of the post-naming phase, and a preference for the semantic distractor at later stages of the trial. These results support a cascaded activation account of lexical access during the first stages of lexical development.

A critical property of bilingual language processing, both in adults and infants, is that of non-selective access to the lexicon. When listening to speech, bilinguals activate all lexical candidates from either language consistent with the speech signal. Even more surprising, when speaking, bilinguals activate all potential words consistent with the concepts they want to express.
This effect has been systematically reported in adults [@marian1999activation; @costa2000cognate; @thierry2007brain; @dunabeitia2009masked].

Also in toddlers [@von2012language; @von2014bilinguals] and school-aged children [@persici2019lexical; @bosma2020cognate]. Relatedly, some recent studies have investigated the recognition of mispronounced words comparing monolingual and bilingual toddlers. Overall, these studies do not find differences between monolinguals and bilinguals [@von2019impact; @wewalaarachchi2017vowels]. Evidence for differential processing of cognate and non-cognate words is provided by Ramon-Casas and colleagues who observed that SpanishCatalan 24-month-olds were more prone to accept mispronunciations when the stimuli involved cognate words than when they involved non-cognate words [@ramon2009vowel; @ramon2010non].

Although recent work has provided strong eviodence of parallel activation in early ages, how cognateness shapes that organisation of the early bilingual lexicon, and how this is reflected on the dynamics of lexical access during word comprehension and production is unclear.

@@

There is evidence of cascaded activation during lexical access. @chow2017spoken adapted the Visual World Paradigm from @huettig2007tug to explore 24- to 30-month-old toddlers' visual fixation patterns during a word recognition when presented with phonological and semantic distractors. In each trial, the authors presented participants with four semantically and phonologically pictures. Four seconds after pictures onset, a word-form was auditorily presented. The word did not refer to any of the pictures displayed on the screen, but was phonologically related to one of them (both labels shared phonological onset), and semantically related to another one of the pictures (both referents belonged to the same taxonomic category). For instance, participants might be presented with the pictures of a sandwich, a bus, a cat, and a dress. Then they would hear the carrier phrase "Look at the **bee**!". The authors registered participants fixations to the phonological and semantic distractors, and found evidence of a preference for the phonological distractor at earlier stages of the post-naming phase, and a preference for the semantic distractor at later stages of the trial. These results support a cascaded activation account of lexical access during the first stages of lexical development.

Evidence from cascaded activation in lexical access is also provided by bilinguals. The bilingual lexicon is language-non selective. There is strong evidence in support of bilinguals activating lexical representations from both languages during both speech production and comprehension. Previous studies on bilingual word production support a cascade account for lexical activation in which the phonology of non-selected lexical representations engages in excitatory or inhibitory dynamics in the lexical selection process.

Bilinguals' trajectories of word recognition  [@vihman2013]

@floccia2020translation tested 27-month-old simultaneous bilinguals in a cross-language priming paradigm. The two languages participants were learning were English, and an Additional Language (either Cantonese, Dutch, Fench, German, Greek, Italian, Mandarin, Polish, Portuguese, or Spanish). A prime word was embedded at the end of a carrier sentence that participants listened to. Then participants were presented with the auditory label of the target word, and two pictures were shown side-by-side, the target picture, and a distractor picture. In Study 1, the target word was the translation equivalent of the prime word (*cheese*-*fromage* for an English-French bilingual). In other trials, prime and target words were unrelated (e.g., *sock*-*fromage*). There was no phonological overlap at onset between prime and target or distractor labels in either of the two conditions. Participants were presented with two blocks of related and unrelated trials. In one of the blocks, participants were presented with English prime words and Home Language target words (*sock*-*fromage*), and in the other block they were presented with Additional Language prime words and English target words (*fromage*-*sock*). The authors found a cross-linguistic priming effect, in which participants showed a stronger preference for the target picture in the post-naming phase in related trials, as compared to unrelated trials. This effect was found regardless of the language of the prime (English or Home Language), or the dominance of the language of the prime (i.e., whether the prime word belonged to the dominant language). Target preference exceeded chance level in both conditions. In Study 2, the authors tested a similar group of bilinguals in a cross-linguistic semantic priing task, in which participants were primed with words in English, and tested with semantically related words in the Additional Language (or vive versa). This time, prime and target words were not translation equivalents, butr rather words whose referents shared semantic features. Results from Study  2 were parallel to those from Study 1. The authors found a strong semantic priming effect accross languages, regardless of the language of the prime or the dominance of the language of the prime. Overall these results suggest that bilinguals recognosed target words in both conditions, and benefitted from a cross-linguistic facilitation effect from translation equivalents. The symmetry of such effect across both languages suggests that participants lexical access ocurred in parallel for the two members of translation equivalents. As @floccia2020translation note, the stimuli set of Study 2 contains multiple cognates, especially in languiages with a larger degree of phonological overlap across translation pairs, like English and Dutch. The authors run *post hoc* analyses exploring the effect of cognateness on the cross-language priming effects found, which showed no evidence of a cognateness effect on target preference. However, the cross-language effects reported in Study 2 might still reflect the confounding influence of cognateness, as the resulting effect might have been generatd by an interaction by the phonological and semantic overlap between cognate primes and targets.

>>>>

targets::tar_load(
	c(model_fits_aggr,
	  model_fits_time,
	  model_loos_aggr,
	  model_loos_time,
	))

@@

This is also the case in the developing lexicon. There is evidence that infants start forming their first lexical representations before their first year of life. It is well established that, across languages, infants start recognising their first word-forms before their first year of life [@jusczyk1999; @halle1994; @vihman2004]. Already at six months of age, infants show evidence of word recognition. When presented with pairs of familiar objects and hearing the label of one of them, 6-months-old infants and older show a looking preference towards the named object [@bergelson20126; @bergelson2015early]. @chow2017spoken adapted the Visual World Paradigm from @huettig2007tug to explore 24- to 30-month-old toddlers' visual fixation patterns during a word recognition when presented with phonological and semantic distractors. In each trial, the authors presented participants with four semantically and phonologically pictures. Four seconds after pictures onset, a word-form was auditorily presented. The word did not refer to any of the pictures displayed on the screen, but was phonologically related to one of them (both labels shared phonological onset), and semantically related to another one of the pictures (both referents belonged to the same taxonomic category. For instance, participants might be presented with the pictures of a sandwich, a bus, a cat, and a dress. Then they would hear the carrier phrase "Look at the **bee**!". The authors registered participants fixations to the phonological and semantic distractors, and found evidence of a preference for the phonological distractor at earlier stages of the post-naming phase, and a preference for the semantic distractor at later stages of the trial. These results support a cascaded activation account of lexical access during the first stages of lexical development.

Evidence from cascaded activation in lexical access is also provided by bilinguals.A critical property of bilingual language processing, both in adults and infants, is that of non-selective access to the lexicon. When listening to speech, bilinguals activate all lexical candidates from either language consistent with the speech signal. Even more surprising, when speaking, bilinguals activate all potential words consistent with the concepts they want to express.
This effect has been systematically reported in adults [@marian1999activation; @costa2000cognate; @thierry2007brain; @dunabeitia2009masked].

Bilinguals infants start recognising words at similar ages. Using a Head-turn Preference Procedure and ERP registration,  @vihman2013 found English-Welsh bilinguals to be sensitive to the prevalence of auditorily presented words. Infants' looking times and ERP differed when presented with words frequently marked as "familiar" by caregivers of demographically equivalent peers (e.g., *nappy*, *thank you*, *bottle*, *apple*, and *naughty*), compared to when presented with rare words (e.g., *Eiffel*, *courtly*, *nettle*, *wacky*, and *juncture*). Bilinguals' trajectories of word recognition  [@vihman2013].

@@

For instance, previous work on lexical acquisition has provided some insight into how bilinguals' vocabulary size is equivalent to monolinguals' across ages, and the possible mechanisms behind this. @hoff2012 found that English-Spanish bilinguals' vocabulary in English was influenced by the words they had already acquired in Spanish. This is consistent with previous accounts of lexical acquisition in which infants acquire new words through preferential acquisition [@hill2013, @fourtassi2020]. A monograph by @floccia2018introduction provided evidence of a language distance effect, where bilinguals learning two languages sharing higher lexical similarity knew more words in their less dominant language than those learning two language with lower lexical similarity. Later work by @mitchell2022cognate and @garcia-castro@cognate provided evidence for a more direct facilitation effect of cognateness. Cognates are form-similar translation equivalents. Cognateness is suggested to facilitate word acquisition through parallel activation: the co-activation of translation equivalents in speech might lead to an faster accumulation of learning instances with words, which ultimately results in an earlier age of acquisition.

Also in toddlers [@von2012language; @von2014bilinguals] and school-aged children [@persici2019lexical; @bosma2020cognate]. Some recent studies have investigated the recognition of mispronounced words comparing monolingual and bilingual toddlers. Overall, these studies do not find differences between monolinguals and bilinguals [@von2019impact; @wewalaarachchi2017vowels]. Evidence for differential processing of cognate and non-cognate words is provided by Ramon-Casas and colleagues who observed that Spanish-Catalan 24-month-olds were more prone to accept mispronunciations when the stimuli involved cognate words than when they involved non-cognate words [@ramon2009vowel; @ramon2010non].

Although recent work has provided strong evidence of parallel activation in early ages, how cognateness shapes that organisation of the early bilingual lexicon, and how this is reflected on the dynamics of lexical access during word comprehension and production is unclear.

@@

@floccia2020translation tested 27-month-old simultaneous bilinguals in a cross-language priming paradigm. The two languages participants were learning were English, and an Additional Language (either Cantonese, Dutch, French, German, Greek, Italian, Mandarin, Polish, Portuguese, or Spanish). A prime word was embedded at the end of a carrier sentence that participants listened to. Then participants were presented with the auditory label of the target word, and two pictures were shown side-by-side, the target picture, and a distractor picture. In Study 1, the target word was the translation equivalent of the prime word (*cheese*-*fromage* for an English-French bilingual). In other trials, prime and target words were unrelated (e.g., *sock*-*fromage*). There was no phonological overlap at onset between prime and target or distractor labels in either of the two conditions. Participants were presented with two blocks of related and unrelated trials. In one of the blocks, participants were presented with English prime words and Home Language target words (*sock*-*fromage*), and in the other block they were presented with Additional Language prime words and English target words (*fromage*-*sock*). The authors found a cross-linguistic priming effect, in which participants showed a stronger preference for the target picture in the post-naming phase in related trials, as compared to unrelated trials. This effect was found regardless of the language of the prime (English or Home Language), or the dominance of the language of the prime (i.e., whether the prime word belonged to the dominant language). Target preference exceeded chance level in both conditions. In Study 2, the authors tested a similar group of bilinguals in a cross-linguistic semantic priming task, in which participants were primed with words in English, and tested with semantically related words in the Additional Language (or vice versa). This time, prime and target words were not translation equivalents, but rather words whose referents shared semantic features. Results from Study  2 were parallel to those from Study 1. The authors found a strong semantic priming effect across languages, regardless of the language of the prime or the dominance of the language of the prime. Overall these results suggest that bilinguals recognised target words in both conditions, and benefited from a cross-linguistic facilitation effect from translation equivalents. The symmetry of such effect across both languages suggests that participants lexical access occurred in parallel for the two members of translation equivalents. As @floccia2020translation note, the stimuli set of Study 2 contains multiple cognates, especially in languages with a larger degree of phonological overlap across translation pairs, like English and Dutch. The authors run *post hoc* analyses exploring the effect of cognateness on the cross-language priming effects found, which showed no evidence of a cognateness effect on target preference. However, the cross-language effects reported in Study 2 might still reflect the confounding influence of cognateness, as the resulting effect might have been generated by an interaction by the phonological and semantic overlap between cognate primes and targets.
